---
  title: "Spectral representation and discretisation"
author: "Finn Lindgren, Liam Llamazares"
date: "September 2021, with major updates May 2022"
output:
  pdf_document: default
html_document: default
bibliography: spde10years-combined.bib
header-includes:
  - \usepackage{times,amsmath,amssymb,gensymb,bm}
- \newcommand{\RR}{\mathbb{R}}
- \newcommand{\ZZ}{\mathbb{Z}}
- \newcommand{\s}{\bm{s}}
- \newcommand{\mmd}{\mathrm{d}}
- \newcommand{\md}{\,\mmd}
- \newcommand{\wt}[1]{\widetilde{#1}}
  - \newcommand{\ol}[1]{\overline{#1}}
    - \newcommand{\wh}[1]{\widehat{#1}}
      - \newcommand{\mc}[1]{\mathcal{#1}}
        - \newcommand{\pN}{\textsf{N}}
        editor_options:
          chunk_output_type: console
        ---

          ```{r setup, include=FALSE}
        knitr::opts_chunk$set(echo = TRUE, cache=FALSE)
        suppressPackageStartupMessages(library(RColorBrewer))
        suppressPackageStartupMessages(library(viridis))
        suppressPackageStartupMessages(library(tidyverse))
        suppressPackageStartupMessages(library(ggplot2))
        suppressPackageStartupMessages(library(patchwork))
        ```
        All of the below discussion is due to Finn Lindgren, the specific spectral distributions chosen in the 2D-example and their plotting are due to Liam Llamazares.

        ## Continuous domains and finite dimensional representations via Fourier theory

        When the correctly folded spectral density for the discrete space process is used, the approximation in the direct spectral representation comes from the frequency resolution, not from frequency truncation.

        The exact spectral representation of the covariance evaluated on a discrete (infinite) lattice is easily derived from the continuous domain representation
        $$
          \begin{aligned}
        R(\s) &= \int_{\RR^d} \exp(i\bm{\omega}\cdot\s) S(\bm{\omega}) \md\bm{\omega} .
        \end{aligned}
        $$
          For simplicity, assume the same lattice spacing $h$ in each direction. The stationary covariance function $R(\s)$ evaluated at lattice points $\bm{k} h$, $k\in\ZZ^d$ is given by
        $$
          \begin{aligned}
        R(\bm{j}h) &= \int_{\RR^d} \exp(i\bm{\omega}\cdot\bm{j} h) S(\bm{\omega}) \md\bm{\omega} \\
        &= \int_{[-\pi/h,\pi/h)^d} \sum_{\bm{k}\in\ZZ^d} \exp(i(\bm{\omega} + 2\pi \bm{k}/h) h \bm{j})
        S(\bm{\omega} + 2\pi \bm{k}/h) \md\bm{\omega}
        \\
        &= \int_{[-\pi/h,\pi/h)^d} \exp(i \bm{\omega} \cdot \bm{j} h)
        \wt{S}(\bm{\omega})
        \md\bm{\omega}, \quad \bm{j}\in \ZZ^d
        \end{aligned}
        $$
          where
        $$
          \begin{aligned}
        \wt{S}(\bm{\omega}) &=\sum_{\bm{k}\in\ZZ^d} S(\bm{\omega} + 2\pi \bm{k}/h),
        \quad \bm{\omega}\in [-\pi/h,\pi/h)^d
\end{aligned}
$$
  If instead the spatial discretisation should be interpreted as the \emph{cell averages} (which is the more usual case for PDE discretisations and e.g. satellite data, rather than pointwise values), the spectrum is altered by a multiplicative frequency
filter with a squared sinc function:
  $$
  \begin{aligned}
\wt{S}(\bm{\omega}) &=\sum_{\bm{k}\in\ZZ^d} S(\bm{\omega} + 2\pi \bm{k}/h)
\prod_{l=1}^d
\left\{
  \frac{
    \sin[(\omega_l + 2\pi k_l/h) h / 2]
  }{
    (\omega_l + 2\pi k_l/h) h / 2
  }
  \right\}^2,
\quad \bm{\omega}\in [-\pi/h,\pi/h)^d .
\end{aligned}
$$


  ### Sampling, in theory

  According to the spectral representation theory for processes with continuous
spectra, we have
$$
  \begin{aligned}
x(\s) &= \int_{\RR^d} \exp(i\bm{\omega}\cdot\s) S(\bm{\omega})^{1/2} \md Z(\bm{\omega}),
\quad \s\in\RR^d ,
\end{aligned}
$$
  where where $\mmd Z(-\bm{\omega})=\ol{\mmd Z(\bm{\omega})}$, $\textsf{Cov}(\mmd Z(\bm{\omega}),\mmd Z(\bm{\omega}'))= \delta(\bm{\omega}-\bm{\omega}')\md\bm{\omega}$ for $\bm{\omega}\in\RR^d$.

                                                                             With the above theory, for sampling, the integral again reduces to a finite domain,
                                                                             but one needs to construct a spectral white noise process
                                                                             $\mmd Z(\bm{\omega})$ with complex conjugate symmetry:
                                                                               \begin{align*}
                                                                             x(\bm{j}h)
                                                                             &= \int_{[-\pi/h,\pi/h)^d} \exp(i \bm{\omega} \cdot \bm{j} h)
                                                                             \wt{S}(\bm{\omega})^{1/2} \md Z(\bm{\omega}), \quad \bm{j}\in \ZZ^d,
                                                                             \end{align*}
                                                                             where $\mmd Z(-\bm{\omega})=\ol{\mmd Z(\bm{\omega})}$, $\textsf{Cov}(\mmd Z(\bm{\omega}),\mmd Z(\bm{\omega}'))= \delta(\bm{\omega}-\bm{\omega}')\md\bm{\omega}$ for $\bm{\omega}\in [-\pi/h,\pi/h)^d$.
                                                                             This can in principle be discretised with a
                                                                             lattice of frequencies in much the same way as a basic integral approximation
                                                                             method that can be used to compute the deterministic integral for the covariance function,
                                                                             with noise variances equal to the cell area/volume of each frequency lattice point.
                                                                             However, the complex conjugacy can be tricky to work with in practice. In the next section
                                                                             we will solve this issue, and provide a fast computational method.

                                                                             ## Practical numerical evaluation with FFT

                                                                             Above we let the spacings $h$ be the same in all dimensions. For greater generality,
                                                                             we now define discrete spacings $\bm{h}=\{h_1,\dots,h_d\}$.
                                                                             Let the number of evaluation points in each dimension be given by
                                                                             $\bm{n}=\{n_1,\dots,n_d\}$. For the cell interpretation of the spatial
                                                                             discretisation, the domain with for the different directions are given by
                                                                             $\bm{L}=\bm{n}\odot\bm{h}$.
                                                                             Define the continuous spectral domain
                                                                             $\mc{I}=\prod_{l=1}^d [-\pi/h_l,\pi/h_l)$, a semi-symmetric index space as
$\mc{K}_0=\prod_{l=1}^d \{-n_l/2,\dots,-1,0,1,2,\dots,n_l/2-1\}$, and
and a non-negative index space
$\mc{K}_+=\prod_{l=1}^d \{0,1,2,\dots,n_l-1\}$, so that if $\bm{k}\in\mc{K}_0$,
then $\bm{k}\bmod \bm{n}\in\mc{K}_+$.

We want to evaluate
$$
  \begin{aligned}
x(\bm{j}\odot\bm{h}) &= \int_{\mc{I}} \exp(i\bm{\omega}\cdot\bm{j} \odot\bm{h}) \wh{x}(\bm{\omega}) \md\bm{\omega}
\end{aligned}
$$
  for all $\bm{j}\in\bm{K}_+$ or all $\bm{j}\in\bm{K}_0$.

The (inverse) FFT in `fftwtools` is a by default unnormalised sum,
$$
  \begin{aligned}
x_{\bm{j}} &= \sum_{\bm{k}\in\mc{K}_+} \exp(i 2\pi\bm{k}\cdot\bm{j} \oslash \bm{n}) \wh{x}_{\bm{k}}
,\quad\bm{j}\in\mc{K}_+ .
\end{aligned}
$$
  so we need to make a Riemann sum approximation of the integral and write it on this form.

Define a discretisation of the $\bm{\omega}$ space via
$$
  (\bm{\omega}_{\bm{k}})_l = \frac{\pi k_l}{h_l n_l/2} = \frac{2\pi k_l}{h_l n_l}, \quad k_l = -n_l/2,\dots,n_l/2-1
$$
  Then
$$
  \begin{aligned}
x(\bm{j}\odot\bm{h}) &\approx \sum_{\{k_l=-n_l/2\}}^{\{n_l/2-1\}} \exp(i2\pi\bm{k}\cdot\bm{j} \oslash \bm{n})
\wh{x}(2\pi\bm{k}\oslash(\bm{n}\odot\bm{h})) \prod_{l=1}^d \frac{2\pi}{h_l n_l}
\end{aligned}
$$
  For the negative $\bm{k}$ indices, we can add a shift by $\bm{n}$ without altering the result (if we also adjust $\wh{x}$ accordingly), since $\exp(i2\pi\bm{n}\cdot\bm{j}\oslash\bm{n})=\exp(i2\pi\bm{j})=1$ for all $\bm{j}$.
We can therefore compute an approximation of the integral by letting
$$
  \wh{x}_{\bm{k} \bmod \bm{n}}=
  \wh{x}(2\pi\bm{k}\oslash(\bm{n}\odot\bm{h}))\prod_{l=1}^d\frac{2\pi}{h_l n_l},
\quad\bm{k}\in\mc{K}_0.
$$


  ### Covariance evaluation

  To compute an approximation of the covariance function, we define
$$
  \sigma_{\bm{k}}^2 = \wt{S}(2\pi\bm{k}\oslash(\bm{n}\odot\bm{h})) \prod_{l=1}^d\frac{2\pi}{h_l n_l},
\quad \bm{k}\in\mc{K}_0,
$$
  and let
$$
  \wh{x}_{\bm{k} \bmod \bm{n}} = \sigma_{\bm{k}}^2,
\quad \bm{k}\in\mc{K}_0 .
$$
  Then $R(\bm{j}\odot\bm{h})\approx x_{\bm{j}\bmod\bm{n}}$ for $\bm{j}\in\mc{K}_0$.

### Process/field simulation

For simulation, the relation $\mmd Z(-\bm{\omega})=\ol{\mmd Z(\bm{\omega})}$ makes
direct construction of the discretised integrand difficult, as we need to keep
track of the complex conjugate symmetry. A practical alternative can be found in
Appendix B.4 of [@LindgrenStocProc], where it is shown that if the complex conjugate
symmetry is ignored, the result of the (inverse) Fourier transform gives
a sample from the process as its real part, and a corresponding sample from the
process _envelope_ as the norms of the absolute values.

Let $\wh{z}_{\bm{k}}$ be iid complex valued $\pN(0,1)$ random variables, and let
$$
  \wh{x}_{\bm{k} \bmod \bm{n}}=
  \wh{z}_{\bm{k}} \sigma_{\bm{k}},
\quad \bm{k}\in\mc{K}_0 .
$$
  Then $x(\bm{j}\odot\bm{h})=\Re[x_{\bm{j}}]$ is an approximate sample from the process model,
and $|x_{\bm{j}}|$ is the corresponding sample of the process envelope, for
$\bm{j}\in\mc{K}_+$.

Check: is there a hidden cyclic boundary condition built in to this method? If so,
is the first half of the process correct?

  ### A note on resolution

  When choosing the frequency resolution for the Fourier integral discretisation,
we are constrained by the desired space resolution $\bm{h}$, but we are
free to choose the number of intervals in each dimension, $\bm{n}$. However,
as a by-product of the FFT construction, this also dictates that the number of evaluation
points in space will _also_ be $\bm{n}$. Thus, even if we only desire to evaluate
the covariance or simulation at a small number of spatial points, we still need
to use enough discretisation points to obtain an accurate Fourier integral approximation.
In practice, this means that we might only use a small portion of the computed results.


## Transformation code

```{r S2C,code=readLines("S2C.R")}
source("S2C.R")
```


## 2D example,

```{r}
#Code below written by Liam Llamazares based on initial version by Finn Lindgren
logkappa <- 0
v1 <- c(0.001,0)
v2 <- c(1,0)
v3 <- c(-1,0)

S_fun1 <- function(omega) S_aniso_matrix(logkappa=logkappa, v=v1, omega=omega)
S_fun2 <- function(omega) S_aniso_matrix(logkappa=logkappa, v=v2, omega=omega)
S_fun3 <- function(omega) S_aniso_matrix(logkappa=logkappa, v=v3, omega=omega)
```
```{r}
n <- c(256, 256)
L <- c(1, 1)
h <- L / n
x_ <- make_x(n, L)
omega_ <- make_omega(n, L)
omega <- as.matrix(expand.grid(omega_))
S_1 <- S_fun1(omega = omega)
S_2 <- S_fun2(omega = omega)
S_3 <- S_fun3(omega = omega)
```

```{r}
S_df <-
  cbind(
    as.data.frame(as.matrix(expand.grid(omega_))),
    data.frame(
      s_1 = S_1,
      s_2 = S_2,
      s_3 = S_3
    )
  ) %>%
  pivot_longer(
    cols = c(s_1, s_2, s_3),
    names_to = "Type", values_to = "Value"
  )

C_df <-
  cbind(
    as.data.frame(as.matrix(expand.grid(x_))),
    data.frame(
      s_1 = as.vector(S2C(S_1, n, h)),
      s_2 = as.vector(S2C(S_2, n, h)),
      s_3 = as.vector(S2C(S_3, n, h))
    )
  ) %>%
  pivot_longer(
    cols = c("s_1", "s_2", "s_3"),
    names_to = "Type", values_to = "Value"
  )
```

```{r}
ggplot(S_df) +
  geom_raster(aes(w1, w2, fill = Value)) +
  scale_fill_gradientn(colours=brewer.pal(7,"Reds"))+
  # scale_fill_viridis(option="inferno")+
  coord_equal()+
  facet_wrap(vars(Type),nrow=1) +
  xlab("omega1") + ylab("omega2") +
  labs(fill = "Spectrum")
ggplot(C_df) +
  geom_raster(aes(x1, x2, fill = Value)) +
  scale_fill_gradientn(colours=brewer.pal(7,"Reds"))+
  coord_equal()+
  facet_wrap(vars(Type),nrow =1) +
  xlab("u1") + ylab("u2") +
  labs(fill = "Covariance")
```




```{r}
S_df %>% group_by(Type) %>% summarise(N = sum(Value > 0))
```
```{r}
n <- c(256, 256)
L <- c(1, 1)
h <- L / n
x_ <- make_x_sampling(n, L)
omega_ <- make_omega_sampling(n, L)
omega <- as.matrix(expand.grid(omega_))
S_1 <- S_fun1(omega = omega)
S_2 <- S_fun2(omega = omega)
S_3 <- S_fun3(omega = omega)
```

```{r}
C_df <-
  cbind(
    as.data.frame(as.matrix(expand.grid(x_))),
    data.frame(
      s_1 = as.vector(S2sample(S_1, n, h)),
      s_2 = as.vector(S2sample(S_2, n, h)),
      s_3 = as.vector(S2sample(S_3, n, h))
    )
  ) %>%
  pivot_longer(
    cols = c("s_1", "s_2", "s_3"),
    names_to = "Type", values_to = "Value"
  )
```

## Sampling example

```{r}
n_sampling <- n * 2
x_ <- make_x_sampling(n_sampling, h)
omega_ <- make_omega_sampling(n_sampling, h)
omega <- as.matrix(expand.grid(omega_))
S_truncated <- S_fun(omega = omega)
S_folded <- fold_spectrum(omega = omega, S_fun = S_fun, h)
S_folded_cell <- fold_spectrum(omega = omega, S_fun = S_fun, h, pointwise = FALSE)
samples_df <-
    cbind(
      as.data.frame(as.matrix(expand.grid(x_))),
      data.frame(
        Truncated = as.vector(S2sample(S_truncated, n_sampling, h, seed = 123L)),
        Folded = as.vector(S2sample(S_folded, n_sampling, h, seed = 123L)),
        "Folded_cell" = as.vector(S2sample(S_folded_cell, n_sampling, h, seed = 123L))
      )
    ) %>%
      pivot_longer(
        cols = c("Truncated", "Folded", "Folded_cell"),
        names_to = "Type", values_to = "Value"
      )
```

```{r}
ggplot(samples_df %>% filter(x1 < L), aes(x1)) +
  geom_line(aes(y = Re(Value), col = Type, lty="Real part")) +
  geom_line(aes(y = Im(Value), col = Type, lty="Imag part")) +
  coord_cartesian(xlim = c(0, L)) +
  xlab("x") + ylab("Sample")
```
(for long correlation range, the three versions are superimposed, since they used the same random seed, and the spectra are almost identical even after folding/truncation)

## References

Note: the theory discussed above follows from the materials found in, among many other places @LindgrenStocProc, ..., and @Cramer1967. The materials in the first two books are later revisions of older teaching materials, and the third is a reprint of the 1967 original:

  * Georg Lindgren, Holger Rootzen, Maria Sandsten (2013),
_Stationary Stochastic Processes for Scientists and Engineers_.
Chapman and Hall
